---
title: 'ECON 421 Project: Some Exploratory Data Analysis'
author: "Derek Situ"
date: "March 9, 2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

Getting the data
```{r message=FALSE, warning=FALSE}
# read in a csv with file path "path", keep only necessary columns and rename
#   for readability, add column "Subtotal" which is equal to the sum of 
#   BestSquat, BestBench, and max(0, Deadlift1, Deadlift2)
read_results <- function(path) {
  read_csv(path, col_names = TRUE) %>%
    select(Name, Sex, Division, 
           WeightClass = "WeightClassKg", Bodyweight = "BodyweightKg", 
           Squat1 = "Squat1Kg", Squat2 = "Squat2Kg", Squat3 = "Squat3Kg", 
           BestSquat = "Best3SquatKg", 
           Bench1 = "Bench1Kg", Bench2 = "Bench2Kg", Bench3 = "Bench3Kg",
           BestBench = "Best3BenchKg",
           Deadlift1 = "Deadlift1Kg", Deadlift2 = "Deadlift2Kg", 
           Deadlift3 = "Deadlift3Kg", BestDeadlift = "Best3DeadliftKg", 
           Total = "TotalKg", Place) %>%
    mutate(Subtotal = BestSquat + BestBench + 
             ifelse(Deadlift1 > 0 | Deadlift2 > 0,
                    ifelse(Deadlift2 > Deadlift1, Deadlift2, Deadlift1),
                    0)) %>%
    relocate(Subtotal, .before = Deadlift3)
}

# read in data
data_21 <- read_results("2021.csv")
data_19 <- read_results("2019.csv")
data_18 <- read_results("2018.csv")
data_17 <- read_results("2017.csv")
data_16 <- read_results("2016.csv")
data_15 <- read_results("2015.csv")
data_14 <- read_results("2014.csv")
data_13 <- read_results("2013.csv")
data_12 <- read_results("2012.csv")
```

What will lifters' beliefs about their opponents' third deadlifts be like? Can we create a regression model to estimate someone's third deadlift as suggested by the TA? Let's check if bodyweight is a good predictor of deadlift capabilities.
```{r message=FALSE, warning=FALSE}
# merge all datasets so that we can run regressions with all of the data
all_data <- rbind(data_21, data_19, data_18, data_17, data_16, data_15,
                  data_14, data_13, data_12) %>%
  select(Name, WeightClass, Bodyweight, Squat3, Deadlift2, Deadlift3) %>%
  mutate(AttSquat3 = abs(Squat3), 
         AttDeadlift2 = abs(Deadlift2), 
         AttDeadlift3 = abs(Deadlift3),
         Percent_Increase = (AttDeadlift3 - AttDeadlift2) / AttDeadlift2,
         D3_Success = ifelse(Deadlift3 > 0, 1, 0)) #%>%
  #filter(!(WeightClass %in% c("84+", "105", "120", "120+"))) ############
```

```{r}
lm_fit <- lm(AttDeadlift3 ~ Bodyweight, data = all_data)
summary(lm_fit)
plot(lm_fit)
```

From the residuals plot, we see that when we estimate Deadlift3 to be 300kg or higher, we are usually overestimating. The next plot shows why:

```{r warning=FALSE}
all_data %>%
  ggplot(aes(x = Bodyweight, y = AttDeadlift3, colour = WeightClass)) +
  geom_point()
```

From plotting bodyweight against attempted Deadlift3, we see that there is a linear relationship!... until bodyweight reaches about 100kg. So when we fit our regression line through the data, it fits nicely until we get to about 100kg, and then we tend to overestimate. It appears that after reaching 100kg, there is no benefit in gaining bodyweight in terms of deadlift. This is a well-known and discussed phenomenon in powerlifting because people who weigh 100kg+ tend to be so large that they cannot physically position themselves to take advantage of optimal leverages and physics. This negative trade-off seems to offset the gains from being heavier at a 1-to-1 ratio.

Okay, we can't easily fit a linear model with bodyweight as a regressor. What about using attempted Squat3 to predict attempted Deadlift3? They both heavily involve the leg muscles.

```{r}
lm_fit <- lm(AttDeadlift3 ~ AttSquat3, data = all_data)
summary(lm_fit)
plot(lm_fit) 
```

Again, whenever we predict large values of attempted Deadlift3 (350kg+), we are overestimating. Let's investigate by looking at the plot of attempted Squat3 vs. attempted Deadlift3:

```{r warning=FALSE}
all_data %>%
  ggplot(aes(x = AttSquat3, y = AttDeadlift3, colour = WeightClass)) +
  geom_point()
```

Similarly to the bodyweight case, there is a linear relationship between attempted Deadlift3 and attempted Squat3... until Squat3 reaches about 300kg. After reaching a Squat3 of 300kg, there seems to be no correlation between a higher squat and a higher deadlift. 

Actually, what we can guess from these two results is that bodyweight is a great predictor of Squat3. Large (100kg+) powerlifters don't have the same problem utilizing optimal leverage in the squat like they do in the deadlift, so squat just increases with bodyweight. Here's a plot to confirm:

```{r warning=FALSE}
all_data %>%
  ggplot(aes(x = Bodyweight, y = AttSquat3, colour = WeightClass)) +
  geom_point()
```

So it seems like we can't easily fit a simple linear model using bodyweight nor squat as regressors. We could take a big non-game-theory-related detour to come up with a more sophisticated prediction procedure that maybe uses a non-linear model...

OR, here's a simple way to predict what a lifter is reasonably capable of doing for Deadlift3, which is and can be actually be used by lifters in a real competition setting: Deadlift2 tends to be about 96% of Deadlift3.
```{r}
mean((all_data$AttDeadlift2 / all_data$AttDeadlift3), na.rm = TRUE)
```
This holds up empirically! In other words, Deadlift3 $\approx$ 1.04Deadlift2. 

All of this whole analysis is just to say that when we model uncertainty of the success of a lifter's Deadlift3, we can consider the fact that typically, lifters think they are reasonably capable of doing 104% of their Deadlift2 for their Deadlift3.




```{r}
# filter for only those with successful Deadlift2
d2_success <- all_data %>% 
  filter(Deadlift2 > 0)

# mean and median ratio (deadlift2 / attempted deadlift3)
mean(d2_success$Deadlift2 / d2_success$AttDeadlift3, na.rm = TRUE) # 0.96
median(d2_success$Deadlift2 / d2_success$AttDeadlift3, na.rm = TRUE) #0.96

# distribution of (deadlift2 / attempted deadlift3) when deadlift3 is successful
# note: Deadlift3 is equal to -n when weight n is attempted and failed
hist(d2_success$Deadlift2 / d2_success$Deadlift3, breaks = 200, 
     main = "Distribution of (Deadlift2/Deadlift3) when both are successful",
     xlab = "Deadlift2 / Deadlift3 (success)",
     xlim = c(0.8, 1))
hist(d2_success$Deadlift2 / d2_success$Deadlift3, breaks = 200, 
     main = "Distribution of (Deadlift2/Deadlift3) when only D2 is successful",
     xlab = "Deadlift2 / Deadlift3 (failed)",
     xlim = c(-1, -0.8))
```





